The MRChem multiresolution analysis code for molecular electronic structure calculations: performance and scaling properties

### 2022-10-26 ReviewersReply.tex

\documentclass[11pt]{article}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{framed}
\usepackage{fullpage}
\usepackage[version=3]{mhchem}
\usepackage[usenames]{xcolor}
\definecolor{darkblue}{rgb}{0.0, 0.0, 0.55}
% for adjustwidth environment
\usepackage[strict]{changepage}
\usepackage{hyperref}

% for formal definitions
\usepackage{framed}

% environment derived from framed.sty: see leftbar environment definition
\definecolor{formalshade}{rgb}{0.95,0.95,1}

\newcommand{\ajwt}[1]{{\color{red}{#1}}}
\newenvironment{formal}{%
  \def\FrameCommand{%
    \hspace{1pt}%
    {\color{darkblue}\vrule width 2pt}%
    {\color{formalshade}\vrule width 4pt}%
    \colorbox{formalshade}%
  }%
  \MakeFramed{\advance\hsize-\width\FrameRestore}%
  \noindent\hspace{-4.55pt}% disable indenting first paragraph
  \begin{adjustwidth}{}{7pt}%
  \vspace{2pt}\vspace{2pt}%
}
{%
  \vspace{2pt}\end{adjustwidth}\endMakeFramed%
}
%\addtolength{\topmargin}{-.6in}
%\addtolength{\textheight}{.2in}
\begin{document}

Dear Editor,\\

We would like to thank the reviewers for their comments on our manuscript.

Yours faithfully,
\\

Peter Wind

Magnar Bj{\o}rgve

Anders Brakestad

Gabriel A. Gerez S.

Stig Rune Jensen

Roberto Di Remigio Eik{\aa}s

Luca Frediani

\clearpage

{\Large \bf Response to Reviewers}

\vspace{1em}

{\bf Editorial Comments}

\begin{formal}
Please include an accessed by date for any reference citing a URL.
\end{formal}

{\bf Reviewer 1}

\begin{formal}
The article "The MRChem multiresolution analysis code for molecular electronic structure calculations: performance and scaling properties" by Wind and coauthors describes implementation and parallel performance of the MRChem program package. The paper is important and should definitely be published, but the presentation is not very concise and can be improved.

To me the article has too many implementation details, that distract from the core message of the article, which would be performance and scaling with system size.
\end{formal}

\begin{formal}
General remarks
 - Section 3 can be omitted entirely, in my opinion it is sufficient to state that the space is adaptively partitioned into 3d nodes/intervals. If necessary this chapter can be shortened and incorporated into Section 4
 \end{formal}

 \begin{formal}
 - I also feel that Sec 2.1 could be better included in Sec. 4
 \end{formal}

\begin{formal}
 - The level of detail is unnecessarily high in Section 4, e.g. explaining the "compressed" and "reconstructed" tree states
 \end{formal}

 \begin{formal}
 - The authors should always make clear what a "node" is: an MRA node (3d interval) or a compute node (possible also an MPI rank?). Also, they write about storing data "together" or "separately" which I assume refers to storing them on the same or different compute nodes? The same is true for statements that data are "locally accessible". This should be clarified.
 \end{formal}

 \begin{formal}
 - Sec 4.2: the naming "MW nodes in real space" confuses me, but I cannot come up with a better suggestion.
 \end{formal}

\begin{formal}
Minor remarks:
 - Sec 2.1: What is the difference between "Product of two MW functions" and "Application of a multiplicative operator"?
 \end{formal}

 \begin{formal}
 - Sec 2.1.1: derivatives are defined in a weak sense, but they are numerically not well-behaved.
 \end{formal}

 \begin{formal}
 - Sec 2.1.2 and 2.1.3: summing and applying does not commute: is there any numerical difference between both approaches?
 \end{formal}

 \begin{formal}
 - Eq. 17: $\sum_p$ instead of $\sum_k$
 \end{formal}

\begin{formal}
Suggestions:
 - The authors state correctly that data sparsity comes with MRA, even if orbitals have non-vanishing contributions from all regions. This is an important statement and has implications for both performance and precision and could be placed more prominently.
\end{formal}

\begin{formal}
 - The authors might mention that many first-order molecular properties, including gradients, can be computed at little extra cost via the Hellmann-Feynman theorem.
\end{formal}

\begin{formal}
 - It would be interesting to see scaling properties with respect to MPI ranks and compute nodes for a given molecule/precision.
\end{formal}

\clearpage

{\bf Reviewer 2}

\begin{formal}
The present manuscript is, according to the authors themselves, a ‘progress report’ to establish that the research ‘is going in the right direction’. Of course, most research publications can be considered as ‘progress reports’ and in the present case, the authors show that their MR implementation of the HF method is now competitive, or even superior, to existing GTO based methods for sufficiently large molecular systems. The manuscript discusses a number of technical points that are essential for achieving the computational performance, and this may be of interest to other groups developing code.
\end{formal}

We thank the reviewer for their nice summary of our work.

\begin{formal}
I find it slightly worrying that two well-established programs, LSDalton and ORCA, using a well-defined method (HF) and analytical integrals, give energy differences in the milli-H range (Table 5) for both the total and atomization energies. As I read the table, this is not due to the use of RI or ADMM, which is the ‘d’-labelled entry. Is it simply a result of a too loose integral screening factor? Perhaps this could be worth a comment.
\end{formal}

We did not investigate the origin of this discrepancy, as our point was to explore the performance of GTO-based and MW-based implementations using their default options. Tightening the integral screening factor and/or the SCF convergence threshold might resolve the discrepancy, but would most likely increase the computational cost.

\end{document}

### 2022-10-26 main.tex
Updates from Overleaf

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{braket}
\usepackage{natbib}
\usepackage{graphicx}

@@

\input{newcommands.tex}

\title{The MRChem MRA code for molecular electronic calculations. Performance and scaling properties}

\author{Peter Wind, Magnar Bjørve, Anders Brakestad, Gabriel Gerez, Stig Rune Jensen, Roberto Di Remigio and Luca Frediani}

%\affilition{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}

@@

\begin{abstract}
MRChem is a Multiresolution Analysis (MRA) code for molecular electronic calculations, based on a multiwavelets adaptive basis representation. A description of the implementation strategy is provided. Benchmark results on systems comprising more than thousand orbitals at the Hartree Fock (HF) level are analyzed, with an emphasis on scaling properties. 
We show that some terms which formally scale as $~0(N^2)$, in effect have a better scaling because of the implicit screening introduced by the inherent adaptivity of the method. Comparison with traditional GTO based software, show that MRChem can be competitive wrt computation time.

\end{abstract}

%test

@@

Gaussian basis sets (GTO and more generally Linear Combination of Atomic Orbitals, LCAO) are well established as a standard for ab initio molecular calculations. 
Even very small basis sets (a few functions per orbital) can give reasonable results for describing molecular properties.

% Husker ikke om jeg skrev dette selv eller om det er tatt fra en artikkel!: In a localized orbital perspective, atom centered basis set are very effective to describe orbitals belonging to that atom, but are basis functions centered on other atoms will be much less useful for describing the local orbitals. For systems with many atoms, one ends up with describing orbitals with a large proportion of very small, coefficients. Those small parts are still necessary among other in order to ensure orthogonality among orbitals. The natural solution is then to describe different orbital or functions in different sets of basis functions. There are many problems associated with this approach, the main one is that each mathematical operation involving more than one function, must relate different basis sets and a new basis has to be defined for every new resulting function. It is a challenge to do this in a balanced and efficient way.

In a MRA framework, the basis can adapt according to the function described (for an in-depth review of the MRA method in the field of Quantum Chemistry, see \citep[]{bischoff2019}). The available basis is in principle infinite, and the basis actually used is dynamically adapted to the required precision. However reasonable results can require the real-space basis to comprise millions of elements for each function. In this sense, the method starts with a big handicap, compared to Gaussian basis sets.

The challenge is not only the large amount of operations needed to perform calculations on large systems, but also the large memory footprint. Using a computer cluster, if not all the data can be stored locally, the data must be stored on separate compute nodes. On modern computers access to data is generally more time consuming than the mathematical operations itself, and if the data is not available locally, this is even more important. The implementation must be able to efficiently copy large amounts of data between compute nodes and the algorithm must be devised so as to use or reuse the data already locally available when possible.

With the exponential development of available computational resources, it has now become possible to perform such MRA calculations on systems comprising thousands of atoms \cite{ratcliff2020}, \cite{madness2016}.

%For larger systems, or high precision, traditional basis set methods require large basis sets, as the same basis is used for all orbitals. Methods exists to "screen" the basis in order to use only a limited part of the entire basis, however the screening will only take effect if the corresponding part of the density matrix can be neglected. Higher accuracy requires larger basis sets, {\sl and} that a larger part of the space must be considered. For those systems the advantages of the Gaussian basis is not obvious.

In this article, we will present the main implementation features of a MRA code, MRChem, which at present is capable of describing systems with thousands of orbitals at the Hartree Fock or Density Functional Theory (DFT) level.

The large memory requirement is addressed by storing the data in a distributed "memory bank", where orbitals can be sent and retrieved independently by any CPU. The total size of the memory bank is then only limited by total size available on the entire computer.

The large computational cost is addressed by parallelization either at the orbital level or through a partition of the real space. This  dual parallelization approach, allows to minimize the relative cost of communication and the overall computational costs.

The multiwavelet approach adapts the precision to the "smoothness" of the potential in a continuous way. The potential resulting from remote subsystems will locally be "smooth" and thereby only require a simplified (coarse grid) description.

Using localized orbitals the adaptiveness of the multiwavelets description will reduce significantly the computational effort required to compute the terms that involves remote orbitals. This opens the way for a linearly scaling method where the linearity arises naturally from the methodology.

Benchmark calculations at the Hartree Fock level show that MRChem is able to describe systems with thousands of electrons, and exhibits near linear scaling properties. Computation times are competitive with state of the art traditional software.

\section{The multiwavelet method for Quantum Chemistry}

Only a superficial overview of the method is given here, as more detailed descriptions can be found elsewhere REF.

The Kohn Sham or HF equations can be written

\begin{equation}
  (-\frac{1}{2}\nabla^2 + V) \psi_i = \sum_{j_{occ}} F_{ij} \psi_j
\end{equation}

\begin{equation}
  F_{ij} = \langle \psi_i |{-\frac{1}{2}\nabla^2+V}| {\psi_j}\rangle
\end{equation}

In the HF case $V=V_{nuc}+J-K$, and in the Kohn Sham equations $V=V_n+J+V_{xc}$, where $V_{nuc}$, $J$, $K$ , $V_{xc}$ are the nuclear, Coulomb, Exchange and exchange-correlation potentials respectively. $F$ is the Fock matrix and $-\frac{1}{2}\nabla^2 = T$ is the kinetic operator.

Compared to more standard LCAO methods, there are a few limitations introduced by the wavelet approach:

Since the  MW functions are not continuous, derivation is problematic. The direct application of the kinetic operator, will not give an accurate result (but good approximations can be found [ref]).

In order to avoid direct use of the Kinetic operator one rewrite the equation as,

\begin{equation}
\psi_i = -2 G_{i}(V\psi_i - \sum_{j_{occ} \ne i} F_{ij} \psi_j)
\end{equation}

where the Helmholtz operator $G_{i}$ is defined by
\begin{equation}
G_{i}(-\nabla^2 - 2 F_{ii} )=1
\end{equation}

Since the number of mw basis is not a priori limited, there is no natural way of defining a set of "virtual orbitals". We cannot define the Fock matrix in the entire basis and even less diagonalize it.

\subsection{Basic scf operations} %put complex conjugates!

We will here show the main steps involved in a SCF iteration in a mw framework.

First the matrix elements of the Fock operator are evaluated explicitly. This involves the application of the kinetic operator ($T$), nuclear  potential ($V_{nuc}$), Coulomb operator ($J$), and  the exchange potential ($K$) and/or the "exchange-correlation" potential from the density functional ($V_{xc}$) on each occupied orbital.

\subsubsection{$T$}
As mentioned the application of the kinetic operator on a mw function can be problematic. However first order derivatives are less problematic than second order derivatives. Further, since the matrix element involves an integration, the resulting numbers are in practice accurate enough. 
In our implementation the matrix elements are computed by first evaluating the first order derivative and the overlap between them:
\begin{equation}
  T_{ij} = -\frac{1}{2} \langle \nabla \psi_i |\nabla {\phi_j}\rangle
\end{equation}

\subsubsection{$V_{nuc}$}

The application of the nuclear potential (and possibly additional external fields) can be performed in two ways. Either the sum over nuclei is performed as a sum of matrix elements for the individual nuclei, or the sum over the potentials is performed first, and the matrix elements evaluated at the end. In our present implementation, the first form is used.
\begin{equation}
  V_{nuc} \phi_j = (\sum_{k = nuclei} V_k) \phi_j
\end{equation}
The total potential for all nuclei ($\sum_{k} V_k$) can be very large in terms of memory. It describes the potential over the entire space, whereas only the parts that overlap with the occupied orbitals are actually used. In the future we plan to adapt the accuracy according to the electronic density.

%The first expression is requires less operations, since only one operator application needs to be evaluated. The second form requires to apply an operator for each nucleus.

%However the second form may still have computational advantages:  The second form is also simpler to set into a parallel framework, since the contribution from each nucleus can be evaluated completely independently. The potential for a single nuclei is memorywise smaller than the total potential.

%I %but we should do something about it to reduce or delocalize the memory footprint

\subsubsection{$J$}

We compute the total density before the total Coulomb potential ($V_C$) is constructed and applied to the orbitals:

\begin{equation}
  V_{C}(r) = \int \frac{\sum_k \phi^*_k(r')\phi_k(r')}{|r-r'|} \ud r'
\end{equation}

\subsubsection{$K$}

The exchange terms are probably the most computationally demanding parts, because they require the application of the Coulomb operator for every orbital pair:

\begin{equation}
  K_{j} = \sum_k  K_{Ex}^k \phi_j 
\end{equation}
with 
\begin{equation}
  K_{Ex}^k(r) \phi_j = \phi_k(r) \int \frac{\phi^*_k(r')\phi_j(r')}{|r-r'|} \ud r'
\end{equation}

Section \ref{Xscreen} provides a more detailed description of how these terms can be evaluated efficiently.

\subsubsection{$V_{xc}$}
In our implementation, the value of the density is evaluated at each point in space, and the dft potential is then evaluated using the external library XCfun[REF], and transformed back into a mw potential.

\subsubsection{Helmholtz operator}

There is one Helmholtz operator associated to each orbital energy, however each Helmholtz operator is applied to a single orbital.

\subsubsection{Basic mw operations required}

The elementary mw operations required for all those steps are
\begin{itemize}
\item Construction of a mw function given an analytic or sampled function.
\item Evaluation of the value of a mw function at a given point in space.
\item Multiplication by a number.
\item Sum of two mw functions.
\item Product of two mw functions.
\item Overlap between two mw functions.
\item Application of a potential on a mw function (multiplication).
\item Evaluation (approximative) of the first order derivative of a mw function.
\item Application of Coulomb operator on a mw function to get a potential.
\item Construction and application of the Helmholtz operator.
\end{itemize}
All these operations are provided by the MRCPP library.

In addition the SCF requires schemes, for orthonormalization, localization and for making the guess of the next orbitals given the preceding iteration results (REF, KAIN).

\subsection{MRChem software, MRCPP and Vampyr}
%section about availability and things the code can do for the user
The MRChem code relies on the MRCPP library for performing the basic mathematical operations in a multiwavelet basis.

git, properties, Vampyr ...

>>>>

\documentclass[journal=jctcce, manuscript=article]{achemso}

\usepackage[utf8]{inputenc}

\setlength {\marginparwidth }{2cm}

\usepackage[table]{xcolor}
\usepackage[version=4]{mhchem}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\graphicspath{{img/}}
 \DeclareGraphicsExtensions{.pdf} %png for low res. pdf for high res. I also have svg as well but other packages needed
\usepackage{caption}
\usepackage{subcaption}
 \DeclareCaptionSubType*[arabic]{figure}
\usepackage{braket}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{todonotes}
\setlength{\columnsep}{1cm}
\usepackage{booktabs}
\usepackage[nohyperlinks,nolist]{acronym}
\usepackage{siunitx}
\usepackage{mathtools}
% to submit to the arXiv:
% 1. uncomment the following
% 2. clear cached files
% 3. recompile from scratch
%\usepackage[newfloat,finalizecache=true,cachedir=.]{minted}
% then:
% 4. comment the line with the finalizecache option
% 5. uncomment the line with the frozencache option
% 6. recompile (to check that it works)
%\usepackage[newfloat,frozencache=true,cachedir=.]{minted}
\usepackage[newfloat]{minted}
%\captionsetup[listing]{position=top}
\setminted{
  fontsize=\footnotesize,
  xleftmargin=\parindent,
  xrightmargin=\parindent,
  breaklines,
  frame=lines
}
\BeforeBeginEnvironment{minted}{\vspace{-7pt}}
\AfterEndEnvironment{minted}{\vspace{-7pt}}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tablefootnote}
\usepackage{multirow}
%\usepackage{hyperref}

\input{acronyms}
\input{newcommands}

\title{The MRChem multiresolution analysis code for molecular electronic structure calculations: performance and scaling properties}

\author{Peter Wind}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\email{peter.wind@uit.no}

\author{Magnar Bjørgve}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}

\author{Anders Brakestad}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}

\author{Gabriel A. Gerez S.}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}

\author{Stig Rune Jensen}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\email{stig.r.jensen@uit.no}

\author{Roberto Di Remigio Eikås}
\affiliation{Algorithmiq Ltd, Kanavakatu 3C, FI-00160 Helsinki, Finland}

\author{Luca Frediani}
\affiliation{Department of Chemistry, UiT - The Arctic University of Norway, N-9037 Tromsø, Norway}
\email{luca.frediani@uit.no}

@@

@@

\begin{abstract}
MRChem is a code for molecular electronic structure calculations, based on a multiwavelet adaptive basis representation. We provide a description of our implementation strategy and several benchmark calculations. Systems comprising more than a thousand orbitals are investigated at the \acl{HF} level of theory, with an emphasis on scaling properties. 
With our design, some terms which formally scale quadratically with the system size, in effect have a better scaling because of the implicit screening introduced by the inherent adaptivity of the method. Comparisons with traditional Gaussian-type orbitals based software, show that MRChem can be competitive with respect to performance.
\end{abstract}

@@

\acp{GTO} and more generally \acp{LCAO}\cite{Jensen.2013.10.1002/wcms.1123} are well established as a standard for \emph{ab initio} molecular electronic structure calculations. As their shape is closely related to the electronic structure of atoms, even very small basis sets with only a few functions per \ac{MO} can give reasonable results for describing molecular properties. However, for extended systems the description of each orbital still requires the contributions from the entire basis in order to ensure orthogonality. Without further precautions, even when using localized orbitals, a large proportion of the coefficients will be very small for those systems.

In a \ac{MRA} framework like \acp{MW}, the basis can adapt according to the function described (for an in-depth review of the \ac{MRA} method in the field of Quantum Chemistry, see Ref.~\citenum{Bischoff2019-mr}). The available basis is in principle infinite and complete, and, in practice, it is dynamically adapted to the local shape of the function and the required precision. This can require the real-space basis to comprise millions of elementary functions for each orbital. In this sense, the method starts with a big handicap compared to \ac{LCAO} basis sets. On the other hand, the exponential growth of available computational resources has in recent years enabled \ac{MRA} calculations on systems comprising thousands of atoms.\cite{Ratcliff2020-hk, Harrison2016-lo}

Two main challenges need to be addressed in order to achieve adequate performance: the large number of operations to perform and the large memory footprint. The former is addressed by limiting the numerical operations to those that are strictly necessary to achieve the requested precision: rigorous error control at each elementary operation is the main strength of a \ac{MW} approach, enabling effective screening. The latter is achieved by algorithmic design: beyond a certain system size, not all data can be stored in local (i.e. fast access) memory. 
On modern computers, data access is generally more time consuming than the mathematical operations, especially if the data is not available locally.\footnote{The online interactive chart at \url{https://colin-scott.github.io/personal_website/research/interactive_latency.html} is useful to understand the relative performance of on-chip \emph{vs.} off-chip memory accesses. In particular, it shows how the evolution of memory chips has been lagging behind that of CPUs.} 
The computer implementation must be able to copy large amounts of data efficiently between compute nodes and the algorithm must be devised to reuse the data already available locally, when possible. In this article, we will present the main implementation features of our \ac{MRA} code, MRChem \cite{mrchem}, to tackle those challenges, thus enabling calculations on large systems at the \ac{HF} level.

Beyond the effective thresholding that screens numerically negligible operations, the large computational cost is addressed by parallelization either at the orbital level or through real space partitioning. This dual strategy allows to minimize the relative cost of communication and the overall computational costs. Further, the most intensive mathematical operations are expressed as matrix multiplications, which allows for optimal efficiency.  
Using localized orbitals, the adaptivity of the \ac{MW} description will significantly reduce the computational effort required to compute the terms involving remote orbitals. This opens the way for a method that scales linearly with system size ($N$), where the linearity arises naturally from the methodology, rather than being obtained with \emph{ad hoc} techniques, such as fast-multipole methods \cite{GREENGARD1987325} for Fock matrix construction, or purification of the density matrix. \cite{Helgaker2000-yb}

The large memory requirement is addressed by storing the data in a distributed ``memory bank", where orbitals can be sent and received independently by any CPU. The total size of the memory bank is then only limited by the overall memory available on the entire computer cluster.
Benchmark calculations at the \ac{HF} level show that MRChem is able to describe systems with thousands of electrons. The implementation exhibits near-linear scaling properties and it can also be competitive with state-of-the-art electronic structure software based on \ac{LCAO} methods.

\section{Solving the Hartree--Fock equations with multiwavelets}

We consider the \ac{SCF} equations of the \ac{HF} method:
\begin{equation}\label{eq:differential-SCF}
  \left(T + V \right) \orbital_i = \sum_{j}^{N_{\mathrm{occ}}} F_{ij} \orbital_j,
\end{equation}
where the Fock matrix elements $F_{ij}$ are obtained as
\begin{equation}\label{eq:fock-matrix-definition}
  F_{ij} = \Braket{ {\orbital_i}|{T+V}|{\orbital_j}}.
\end{equation}
In the above equations, $T = -\frac{1}{2}\nabla^2$ is the kinetic
energy operator, and the potential $V$ contains the nuclear-electron
attraction $V_{\mathrm{nuc}}$, the Coulomb repulsion $J$, and the exact exchange $K$. 
To solve the \ac{SCF} equations within a \ac{MW} framework, we rewrite the differential equation \eqref{eq:differential-SCF} as an integral equation:
\begin{equation}\label{eq:integral-SCF}
\orbital_i = -2 H^{\mu_i}\star\left[V\orbital_i - \sum_{j \ne i}^{N_{\mathrm{occ}}} F_{ij} \orbital_j\right],
\end{equation}
where $\star$ stands for the convolution product.
This form is obtained by recalling that the shifted kinetic energy operator $T-\epsilon$ has a closed-form Green's function:\cite{Kalos1962-ok,Beylkin2005-kg}
\begin{equation}\label{eq:helmholtz-green}
    (T-\epsilon)^{-1} = 2 H^{\mu}, \quad H^{\mu} = \frac{e^{-\mu r}}{r}, \quad \mu = \sqrt{-2\epsilon},
\end{equation}
By setting $\epsilon_i = F_{ii}$ and $\mu_i = \sqrt{-2F_{ii}}$, the diagonal term is excluded from the summation in Eq.~\eqref{eq:integral-SCF}.

A \ac{MW} representation does not have a fixed basis. It is therefore not possible to express functions and operators as vectors and matrices of a predefined size, and the virtual space is to be regarded as infinite. Still, each function has a finite, truncated representation defined by the chosen precision threshold, but this representation will in general be different for different functions. It is therefore necessary to develop working equations which allow the direct optimization of orbitals. On the other hand, only occupied orbitals are constructed and the coupling through the Fock matrix in Eqs. \eqref{eq:differential-SCF} and \eqref{eq:integral-SCF} is therefore limited in size.

Differential operators such as the Laplacian pose a fundamental problem for function representations which are not continuous, and a naive implementation leads to numerical instabilities. Robust approximations are nowadays available\cite{Anderson2019-bx}, but for repeated iterations avoiding the use of the Laplacian operator is still an advantage. This is the main reason for using the integral form in Eq.~\eqref{eq:helmholtz-green} rather than the differential form.

Additionally, it can be shown that iterating on the integral equation corresponds to a preconditioned steepest descent. Practical applications show that this approach is comparable in efficiency (as measured in the rate of convergence of the \ac{SCF} iterations) to more traditional methods.\cite{Harrison2004-fs}

\subsection{The basic operations in the SCF iterations}

We will here list the main operations involved in the \ac{SCF} procedure, with emphasis on those aspects which are relevant for an efficient parallelization. For a detailed description of the \ac{SCF} optimization using \acp{MW}, the reader is referred to a recently published article which describes the algorithm in detail.\cite{Jensen:2022gg}

The starting point is an initial set of orthonormal orbitals which define a Slater determinant for the given problem. A density is computed and the corresponding operators ($J$ and $K$) are built. The Fock matrix is computed as in Eq.~\eqref{eq:fock-matrix-definition}, which requires the application of the kinetic operator ($T$), nuclear potential ($V_{\mathrm{nuc}}$), Coulomb ($J$) and exchange ($K$) operators on each occupied orbital. The argument for the application of the integral operator is assembled (the square bracket in Eq.~\eqref{eq:integral-SCF}). Then the integral operator is applied separately for each orbital and the cycle is iterated after an orthonormalization step. Algorithm \ref{alg:SCF-iter} depicts these main steps.

\begin{algorithm*}
  \caption{Pseudocode for a single self-consistent field iteration. Each step takes advantage of adaptivity and screening, as described in Sections~\ref{sec:localvsorb} and~\ref{sec:adaptivity-screening}.} 
  \label{alg:SCF-iter}
  \begin{algorithmic}[1]
    \Procedure{SCF iteration}{$\lbrace \orbital_{i}^{[n]}\rbrace_{i=1}^{N_{\mathrm{occ}}}$, $\epsilon$}

    \State Build Coulomb operator: $$J^{[n]} =
    \sum_i \left[\frac{1}{|\rvec - \rvec^\prime|} \star\left(\orbital_{i}^{*,[n]}\orbital_{i}^{[n]}\right)\right]$$
    \Comment{See Section~\ref{sec:compute-J}}
    \State Build and apply Exchange operator: $$K^{[n]}\orbital_{j}^{[n]} = \sum_i \orbital_{i}^{[n]} \left[\frac{1}{|\rvec - \rvec^\prime|} \star\left(\orbital_{i}^{*,[n]}\orbital_{j}^{[n]}\right)\right]$$
    \Comment{See Section~\ref{sec:compute-K}}
    \State Assemble Fock matrix elements: $$F_{ij}^{[n]} = \Braket{\orbital_{i}^{[n]} | T + V_{\mathrm{nuc}} + J^{[n]} - K^{[n]}| \orbital_{j}^{[n]}}$$
    \Comment{See Sections~\ref{sec:compute-T} and~\ref{sec:compute-V}}
    \State Assemble right-hand side: $$V^{[n]}\orbital_{i}^{[n]} - \sum_{j\neq i}^{N_{\mathrm{occ}}}F_{ij}^{[n]}\orbital_{j}^{[n]}$$
    \Comment{See Section~\ref{sec:rhs}}
    \State Update orbitals: $$\orbital_{i}^{[n+1]} = -2H^{\mu_{i}}\star\left[ V^{[n]}\orbital_{i}^{[n]} - \sum_{j\neq i}^{N_{\mathrm{occ}}}F_{ij}^{[n]}\orbital_{j}^{[n]}\right]$$
    \Comment{See Section~\ref{sec:rhs}}
    \State Accelerator and localization: A new set of orbitals is created using results from previous iterations and localization. This step is optional and does not need to be performed at each iteration.
    \EndProcedure
  \end{algorithmic}
\end{algorithm*}

The elementary mathematical operations required in each SCF iteration are:
\begin{itemize}
\item Projection of functions in a \ac{MW} representation, given their analytic form or a representation on an arbitrary grid.\footnote{As long as a function is \emph{algorithmically} known, that is, there exists an implementation that given a point in the domain, for example $\mathbb{R}^{3}$, returns the function value, then it is possible to obtain its \ac{MW} representation.} 
\item Evaluation of the \ac{MW} representation of functions at a given point in space.
\item Multiplication of a function by a number.
\item Sum of two \ac{MW} functions.
\item Product of two \ac{MW} functions.
\item Computation of the overlap integral between two \ac{MW} functions.
\item Application of a multiplicative operator, \emph{e.g.} a potential, on a \ac{MW} function.
\item Evaluation of the first order derivative of a \ac{MW} function.
\item Construction of the separated representation of the Poisson and Helmholtz kernels as \ac{MW} functions.
\item Application of the Poisson or Helmholtz operators on a \ac{MW} function.
\end{itemize}

All these operations are provided by the MRCPP library\cite{mrcpp}. They must be composed to achieve a \ac{SCF} implementation, and they need to be integrated in a parallel framework to achieve good scaling and performance. In addition, the \ac{SCF} cycle requires schemes for orbital orthonormalization, rotation, and localization, Fock matrix diagonalization, history storage and utilization to enable convergence acceleration, using \emph{e.g.}~the \ac{KAIN} method.~\cite{Harrison2004-fs}

\subsubsection{The kinetic operator $T$}
\label{sec:compute-T}
The application of the kinetic energy operator on a \ac{MW} function is formally not well defined due to the discontinuities in the basis.\cite{Anderson2019-bx} However, when adopting an integral reformulation of the \ac{SCF} equations, the kinetic energy operator is required only in the  expectation value expressions as in Eq.~\eqref{eq:fock-matrix-definition}, and as shown by \citeauthor{Anderson2019-bx}~\cite{Anderson2019-bx} the original definition of the first-order derivative by \citeauthor{Alpert:2002cx}~\cite{Alpert:2002cx} is able to deliver sufficient precision in this specific case. On the basis of this consideration, and for computational efficiency, the matrix elements of the Laplacian operator are computed as inner products of orbital gradients:
\begin{equation}
  \label{eq:kinetic-matrix}
  T_{ij} = 
  -\frac{1}{2} 
  \Braket{\orbital_i | \nabla^2 \orbital_j} 
  = 
  \frac{1}{2} 
  \Braket{\nabla \orbital_i | \nabla \orbital_j}.
\end{equation}

It is possible to avoid the kinetic energy operator altogether \cite{Jensen:2022gg}, but in practice this complicates the algorithm to the point where the direct computation as in Eq.~\eqref{eq:kinetic-matrix} is to be preferred for large systems.

\subsubsection{The nuclear potential $V_{\mathrm{nuc}}$}
\label{sec:compute-V}
The nuclear potential consists of a sum of contributions from the different nuclei. As such, applying it to a given orbital can be performed in two ways.

\begin{description}
\item[apply-then-sum] The potential from each nucleus $\alpha$ is applied to each orbital and the results summed up over all nuclei:
\begin{equation}
  V_{\mathrm{nuc}} \orbital_j = \sum_{\alpha}^{N_{\mathrm{atoms}}} \left(V_{\alpha} \orbital_j\right),
\end{equation} 
\item[sum-then-apply] The sum over the nuclei is performed first, followed by the multiplication with the given orbital:
\begin{equation}
V_{\mathrm{nuc}} \orbital_j = \left(\sum_{\alpha}^{N_{\mathrm{atoms}}} V_{\alpha}\right) \orbital_j.
\end{equation}
\end{description}

The first approach requires $N_{\mathrm{atoms}}\times N_{\mathrm{occ}}$ function multiplications, followed by a summation of $N_{\mathrm{atoms}}$ terms. It also needs memory for $N_{\mathrm{atoms}}\times N_{\mathrm{occ}}$ \ac{MW} functions. However, since these objects are relatively localized, their memory footprint is modest and can, in addition, be amortized by distributing objects across compute nodes. Furthermore, the function-function products can also be distributed across nodes and screened aggressively.

The second approach requires a summation of $N_{\mathrm{atoms}}$ terms, followed by $N_{\mathrm{occ}}$ function multiplications. However, the total potential is no longer well-localized, as it describes the potential from all clamped nuclei, and it requires a large amount of memory storage. Storing the nuclear potential is wasteful: only its \ac{MW} nodes with non-zero overlap with the occupied orbitals will contribute to the final result.

Despite these considerations, our present implementation uses the ``sum-then-apply" form. In the future we plan to try the first approach and also adapt the accuracy according to the electronic density and compare the performance between the two approaches.

\subsubsection{The Coulomb potential $J$}\label{sec:compute-J}
The Coulomb interaction can be expressed as:
\begin{equation}
    \begin{aligned}
  J(\rvec) &= \sum_i \int \frac{\orbital^*_i(\rvec')\orbital_i(\rvec')}{|\rvec-\rvec'|} \, \mathrm{d}\rvec' =
  \sum_i  \left[
  \frac{1}{|\rvec-\rvec'|} \star 
  \orbital^*_i\orbital_i
  \right] =
  \frac{1}{|\rvec-\rvec'|} \star 
  \left[\sum_i 
  \orbital^*_i\orbital_i
  \right]
    \end{aligned}
\end{equation}
where, as indicated by the last step, we can choose the order of the operations when constructing the Coulomb potential: either summing the orbital densities first and compute the potential of the total density (\textbf{sum-then-apply}), or, first computing the Coulomb potential of each individual orbital density and then sum the potentials (\textbf{apply-then-sum}). The electrostatic potential is in general a much smoother function than its corresponding density, which contains cusps at every nuclear site. It is thus desirable to avoid the explicit construction of the global density and instead gradually build a global potential by partial applications of the Poisson operator to subsets of orbital densities. This operation can also easily be distributed in parallel by assigning different subsets to different processes.

The Poisson kernel is expressed in a separable form, within the requested precision, as a sum of Gaussian contributions.\cite{Beylkin.10.1016/j.acha.2005.01.003} Each Gaussian kernel is applied separately and the contributions are then added up. The separated representation is essential to achieve fast linear-scaling algorithms: on the one hand it brings down the formal scaling of the operator application for each node from $2^6 k^6$ to $M 2^4 k^4$, where $k$ is the polynomial order and $M$ is the separation rank; on the other hand the Gaussian contributions will range from wide ones (shallow \ac{MW} representation which couple long-range but at coarse scales) to narrow ones (deep \ac{MW} representation which couple short-range at all scales). In practice, most of the $M$ terms will be applied very efficiently and the separation between long-range and short-range interactions -- which is otherwise achieved through the \ac{FMM} -- is naturally included without extra effort.\cite{Frediani.10.1080/00268976.2013.810793, Beylkin.10.1016/j.acha.2007.01.001}

Normally the potential is computed with the required precision in the entire space, and since it is long range, it will be relatively memory consuming.

\subsubsection{The exchange potential $K$}\label{sec:compute-K}
The construction and application of the exchange potential is the most computationally demanding part, because the exchange operator cannot be written as a simple multiplicative potential. A Coulomb potential must be constructed and applied for each orbital pair:
